% ===========================================================================
%  Nonlinear Algebra -- fill-in notes on Chapters 1-2
%  (Michalek & Sturmfels, "Invitation to Nonlinear Algebra", AMS GSM 211)
%  woven on LOOM.   compile:  latexmk -xelatex main   (or xelatex twice)
%
%  HOW THESE NOTES WORK
%   - Read the strands and the stated theorems (passive intake).
%   - Fill the blanks \fillin, the \TODO proof gaps, and the "your turn" boxes
%     (active recall). The book is your answer key.
% ===========================================================================
\documentclass[cjk]{loom}   % [cjk] only for the pen name; all prose is English

% --- pedagogy tools (\block \trigger \TODO \fillin \recall yourturn \workspace)
%     and booktabs/tabularx/L now come built into loom.cls v0.2 ---------------

% --- math shorthands --------------------------------------------------------
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\V}{\mathcal{V}}        % variety  V(I)
\newcommand{\Ideal}{\mathcal{I}}    % vanishing ideal  I(W)
\newcommand{\gen}[1]{\langle #1\rangle}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\inn}{\operatorname{in}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\rad}{rad}
\DeclareMathOperator{\HS}{HS}
\DeclareMathOperator{\HP}{HP}
\DeclareMathOperator{\rank}{rank}

\runningthread{nonlinear algebra · ch 1--2}

\begin{document}

\loomcover
  {Nonlinear Algebra}
  {Fill-in notes · Chapters 1--2 · Michałek \& Sturmfels}
  {北极甜虾 (剑心犹在！)}
  {\today}

% ===========================================================================
\section*{The dictionary, and how to read these notes}
\addcontentsline{toc}{section}{The dictionary · how to read}
\warmth{0}

\begin{strand}
The whole of Chapters 1--2 is \emph{one} idea, stated twice by Sophie Germain:
\keyword{``Algebra is but written geometry''} (Ch.\,1) and
\keyword{``Geometry is but drawn algebra''} (Ch.\,2). There is a dictionary
between the \emph{algebra} side (ideals in the polynomial ring $K[\xb]$) and the
\emph{geometry} side (varieties in $K^n$ or $\PP^n$). Learn the dictionary and
the two chapters become one.
\end{strand}

\block{How to read these notes (passive \emph{and} active)}
\begin{itemize}[leftmargin=1.3em,itemsep=1pt]
  \item \keyword{Read} the strands and the stated theorems. That is the passive intake.
  \item \keyword{Fill} the blanks \fillin[1.2cm], the \TODO{the missing step} gaps inside proofs,
        and the \emph{Your turn} boxes. That is the active recall. The book is your answer key.
  \item Bump each section's opening \texttt{\textbackslash warmth\{0\}} to $0$--$5$ once you can rebuild it cold.
\end{itemize}

\block{The master dictionary (fill it in as you go)}
\noindent\begin{tabularx}{\linewidth}{@{}L L@{}}
\toprule
{\color{indigo}Algebra: ideals in $K[\xb]$} & {\color{madder}Geometry: varieties in $K^n,\PP^n$}\\
\midrule
ideal $I$                       & variety $\V(I)$ \;(§\ref{sec:affine})\\
vanishing ideal $\Ideal(W)$ (radical) & variety $W=\V(\Ideal(W))$\\
$I\subseteq J$                  & $\V(I)\;\fillin[1cm]\;\V(J)$ \;(which way?)\\
sum $I+J$                       & $\fillin[2cm]$ of $\V(I),\V(J)$\\
intersection $I\cap J$          & $\fillin[2cm]$ of $\V(I),\V(J)$\\
prime ideal                     & \fillin[3cm] variety \;(§\ref{sec:affine})\\
maximal ideal $\gen{x_1-p_1,\dots,x_n-p_n}$ & a single point $\fillin[1.5cm]$\\
radical ideal                   & ($\Ideal\!=\!\V$ inverse, up to radical: Nullstellensatz, Ch.\,6)\\
quotient $K[\xb]/\Ideal(W)$     & coordinate ring $K[W]$ = functions on $W$\\
dimension of $I$ (§\ref{sec:dimdeg}) & dimension of $\V(I)$\\
degree of $I$                   & \# points of a general linear slice\\
$\gen{f,\partial f/\partial x_1,\dots}$ has no zeros & $\V(f)$ is \fillin[2cm]\\
homogeneous ideal               & projective variety \;(§\ref{sec:proj})\\
primary decomposition (Ch.\,3)  & decomposition into irreducibles\\
\bottomrule
\end{tabularx}

\begin{strand}
The map $\V$ sends algebra to geometry; the map $\Ideal$ sends geometry back to
algebra. They are \emph{inclusion-reversing}, and they are almost inverse to each
other. ``Almost'' is the content of the Nullstellensatz (Chapter 6); for now,
$\Ideal(\V(I))=\rad(I)$ is the slogan to carry.
\end{strand}

\newpage
{\small\tableofcontents}

\clearpage
% ===========================================================================
\input{sections/01-ideals}
\input{sections/02-groebner}
\input{sections/03-dimension-degree}
\input{sections/04-affine-varieties}
\input{sections/05-projective-varieties}
\input{sections/06-low-dimensions}

\end{document}
