% ===========================================================================
\section{The best simple picture}\label{sec:eckart}
\warmth{0}\quad\whisper{Keep the top few singular triples and you have the best low-rank copy of $A$. Provably.}

\begin{strand}
Here is why the SVD is everywhere in data science. Write $A$ as a sum of rank-one layers,
\[
  A=\sum_{i=1}^{r}\sigma_i\,u_i v_i\T ,
\]
ordered by importance $\sigma_1\ge\sigma_2\ge\cdots$. Keep only the top $k$ layers and you get
$A_k=\sum_{i\le k}\sigma_i u_i v_i\T$, the \keyword{truncated SVD}. The miracle: among \emph{all}
matrices of rank $\le k$, $A_k$ is the closest one to $A$. Compression, denoising, PCA,
recommender systems, are all this one theorem.
\end{strand}

\begin{theorem}[Eckart--Young--Mirsky]\label{thm:ey}
For every $B$ with $\rank(B)\le k$,
\[
  \norm{A-B}\;\ge\;\norm{A-A_k},
\]
in both the operator and the Frobenius norm. The optimal error is read off the discarded
singular values:
\[
  \norm{A-A_k}_2=\sigma_{k+1},\qquad
  \norm{A-A_k}_F=\sqrt{\sigma_{k+1}^2+\cdots+\sigma_r^2}.
\]
\end{theorem}
\begin{proof}
Two halves:
\begin{enumerate}[leftmargin=2em,itemsep=1pt]
  \item \emph{The error of $A_k$.} $A-A_k=\sum_{i>k}\sigma_i u_i v_i\T$ is itself in SVD form, so
        its singular values are $\sigma_{k+1},\sigma_{k+2},\dots$. \TODO{read $\norm{\cdot}_2$ and $\norm{\cdot}_F$ off those}
  \item \emph{Nobody does better.} Let $\rank(B)\le k$, so $\Null(B)$ has dimension $\ge n-k$.
        It meets $\operatorname{span}\{v_1,\dots,v_{k+1}\}$ (dimension $k+1$) in some unit vector $w$.
        \TODO{show $Bw=0$ and $\norm{Aw}\ge\sigma_{k+1}$, hence $\norm{A-B}_2\ge\norm{(A-B)w}=\norm{Aw}\ge\sigma_{k+1}$}
\end{enumerate}
\end{proof}

\begin{strand}
The reason this works is the ordering. Total ``energy'' $\norm{A}_F^2=\sum_i\sigma_i^2$ piles up in
the largest few singular values, so throwing away the small ones costs almost nothing. A picture,
a covariance matrix, a ratings table: each is usually \emph{approximately} low rank, meaning its
$\sigma_i$ decay fast, and the truncated SVD is a faithful thumbnail at a fraction of the storage.
\end{strand}

\begin{yourturn}
Suppose $A$ ($4\times4$) has singular values $\sigma=(10,\,6,\,2,\,1)$. For the rank-$2$
truncation $A_2$:
\[
  \norm{A-A_2}_2=\fillin[1cm],\qquad
  \norm{A-A_2}_F=\fillin[2cm],\qquad
  \frac{\norm{A_2}_F^2}{\norm{A}_F^2}=\fillin[2.2cm].
\]
\recall{Spectral error $=\sigma_3=2$; Frobenius error $=\sqrt{2^2+1^2}=\sqrt5$; energy kept $=\frac{10^2+6^2}{10^2+6^2+2^2+1^2}=\frac{136}{141}\approx 96.5\%$. Two numbers carry the matrix.}
\end{yourturn}

\begin{strand}
Caveat worth knowing: Eckart--Young is for \emph{unitarily invariant} norms (operator, Frobenius,
all Schatten-$p$). For other measures (entrywise $\ell_1$, nonnegative or sparse factors), best
low-rank approximation can be NP-hard. The SVD's optimality is special, not universal.
\end{strand}
