% ===========================================================================
\section{The Cramér–Chernoff engine}\label{sec:chernoff}
\warmth{0}\quad\whisper{The trunk. Every later engine just works on one step of this pipeline.}

\begin{strand}
In one line: to bound the probability that a random quantity strays far from its
mean by something \keyword{exponentially small}, exponentiate the tail through the
Laplace transform (the MGF) and pick a good $\lambda$ to push the right-hand side
down. The whole subject is variations on this single move.
\end{strand}

\block{The paradigm}
\trigger{You want an exponentially small bound on $\Pr[X\ge a]$, with the MGF of $X$ under control.}
For every $\lambda>0$,
\[
  \Pr[X\ge a]=\Pr\!\big[e^{\lambda X}\ge e^{\lambda a}\big]\le \E[e^{\lambda X}]\,e^{-\lambda a}
  \qquad(\text{Markov on }e^{\lambda X}),
\]
then take the infimum over $\lambda$. Writing the cumulant generating function
$\psi_X(\lambda)=\log\E[e^{\lambda X}]$,\warp{mgf}
\[
  \log\Pr[X\ge a]\le -\psi_X^*(a),\qquad
  \psi_X^*(a)=\sup_{\lambda>0}\big(\lambda a-\psi_X(\lambda)\big),
\]
where $\psi_X^*$ is the \keyword{Legendre transform} (the rate function) of $\psi_X$.

\begin{strand}
Three-line heuristic: (1) independence $\Rightarrow$ MGFs \emph{multiply}, which is the
\emph{only} reason a ``sum'' gets an exponential bound; (2) \emph{center} first,
$X\leftarrow X-\E X$, for the cleanest argument; (3) bound each $\E[e^{\lambda X_i}]$ by one
\emph{termwise convex} estimate. Every theorem in the subject just swaps that estimate
and that $\lambda$.
\end{strand}

\block{Cheat-sheet: the three opening moves}
\noindent\begin{tabularx}{\linewidth}{@{}l l L@{}}
\toprule
{\color{indigo}Tool} & {\color{indigo}Setting} & \multicolumn{1}{l}{\color{indigo}Tail bound / when to use}\\
\midrule
Markov           & $Y\ge 0$                     & $\Pr[Y>\alpha\E Y]<1/\alpha$; opening move, first-order info only\\
Chebyshev        & $\Var X<\infty$              & $\Pr[|X-\E X|>t]\le \Var X/t^2$; second-order info\\
Chernoff (master)& MGF exists                   & $\Pr[X\ge a]\le e^{-\psi_X^*(a)}$; want an exponential tail\\
\bottomrule
\end{tabularx}

\block{Main results \& proof skeletons}

\begin{theorem}[Chernoff master bound]\label{thm:chernoff-master}
If the MGF of $X$ is finite near $0$, then $\displaystyle \Pr[X\ge a]\le e^{-\psi_X^*(a)}$.
\end{theorem}
\begin{proof}
Skeleton: \emph{(i)} Markov on $e^{\lambda X}$; \emph{(ii)} take logs to get $\log\Pr[X\ge a]\le \psi_X(\lambda)-\lambda a$;
\emph{(iii)} take the infimum over $\lambda>0$. \TODO{add convexity of $\psi_X$, the attaining $\lambda$, and the equality case}
\end{proof}

\begin{lemma}[Additivity: independence makes $\psi$ add]\label{lem:mgf-add}
$X_1,\dots,X_n$ independent $\Rightarrow \psi_{\sum_i X_i}=\sum_i \psi_{X_i}$.
\end{lemma}
\begin{proof}\TODO{one line: $\E[e^{\lambda\sum X_i}]=\prod\E[e^{\lambda X_i}]$, take logs}\end{proof}

\begin{example}[The sub-Gaussian seed: a symmetric $\pm1$ sum]\label{ex:rademacher}
Let $\varepsilon_i$ be independent uniform $\pm1$ and $S_n=\sum_{i\le n}\varepsilon_i$. From
$\cosh\lambda\le e^{\lambda^2/2}$ we get $\psi_{S_n}(\lambda)\le \lambda^2 n/2$, and $\lambda=a/n$ gives
\[
  \Pr[S_n\ge a]\le e^{-a^2/2n},\qquad \Pr[|S_n|\ge a]\le 2e^{-a^2/2n}.
\]
\TODO{prove $\cosh\lambda\le e^{\lambda^2/2}$ by termwise Taylor comparison; this is the seed inequality of the whole subject}
\end{example}

\begin{strand}
Why it goes this way: $\cosh\lambda\le e^{\lambda^2/2}$ is the \emph{seed inequality}. Replace it by a
different termwise estimate (Hoeffding's lemma $\to$ bounded variables, Bennett/Bernstein
$\to$ variance-aware) and you grow §\ref{sec:subgaussian}–§\ref{sec:entropy}. Swap the estimate, swap $\lambda$, that is all.
\end{strand}

\block{Tightness: Chernoff gets the large-deviation rate right}
\begin{theorem}[Cramér, matching lower bound]\label{thm:cramer}
Under mild conditions $\displaystyle \lim_{n\to\infty}\tfrac1n\log\Pr\!\big[\tfrac1n\textstyle\sum_i X_i\ge a\big]=-\psi_{X_1}^*(a)$.
\end{theorem}
\begin{proof}
Skeleton: the upper bound is Thm~\ref{thm:chernoff-master}+Lem~\ref{lem:mgf-add}; the lower bound uses
\emph{exponential tilting} (change of measure) to make $a$ the mean under the new measure, then CLT.
\TODO{construct the tilted measure $d\mathbb{Q}\propto e^{\lambda x}\,d\mathbb{P}$}
\end{proof}

\loose{The three deviation regimes (keep this picture): small deviations $\to$ normal; large deviations $\lambda_n=o(\sigma_n)\to \log\Pr\sim-\lambda_n^2/2$; very large deviations $\lambda_n=\Omega(\sigma_n)\to$ normal fails, Poisson takes over. Cf.\ Appendix A.}%
\pick{mgf} This thread is upgraded in §\ref{sec:subgaussian} to ``sub-Gaussian $\equiv$ MGF dominated by $e^{\sigma^2\lambda^2/2}$''.
