% ===========================================================================
\section{Talagrand's convex distance}\label{sec:talagrand}
\warmth{0}\quad\whisper{A product space has no geometry, only convexity. Talagrand's convex distance adapts to ``how many bad coordinates''.}

\begin{strand}
In one line: on a pure \emph{product} probability space (no sphere-like geometry), Talagrand's
\keyword{convex distance} yields dimension-free, very strong concentration for convex Lipschitz functions.
Its edge over bounded differences (§\ref{sec:bdd}): it does not look at the ``largest single-coordinate jump''
but at ``how \emph{many} coordinates you must change to escape a set'', adapting to sparse dependence.
\end{strand}

\block{The paradigm}
\trigger{$f$ on a product space is \keyword{convex} Lipschitz, or escaping an event $A$ requires changing many coordinates.}
Define the \keyword{convex distance} $d_T(x,A)=\sup_{\|\alpha\|_2\le1}\ \inf_{y\in A}\sum_{i:x_i\ne y_i}\alpha_i$ (weighted Hamming).
Talagrand: $\E\big[e^{d_T(X,A)^2/4}\big]\le 1/\Pr[A]$, hence
\[
  \Pr[A]\cdot\Pr[d_T(X,A)\ge t]\le e^{-t^2/4}.
\]

\begin{strand}
Three-line heuristic: (1) $d_T$ is a min-max of \emph{weighted} Hamming, automatically finding the ``cheapest escape direction'';
(2) once $\Pr[A]\ge\tfrac12$, the convex distance from $A$ concentrates sub-Gaussianly;
(3) for a \emph{convex} $L$-Lipschitz $f$, take $A=\{f\le\operatorname{Med}f\}$ to get $\Pr[|f-\operatorname{Med}f|>t]\le 4e^{-t^2/4L^2}$, dimension-free.
\end{strand}

\block{Cheat-sheet: convex distance and its corollaries}
\noindent\begin{tabularx}{\linewidth}{@{}l l L@{}}
\toprule
{\color{indigo}Result} & {\color{indigo}Setting} & \multicolumn{1}{l}{\color{indigo}Statement / when to use}\\
\midrule
Convex-distance ineq.\ & product space, any $A$    & $\Pr[A]\Pr[d_T\ge t]\le e^{-t^2/4}$; the master switch\\
Convex Lipschitz    & $f$ convex, $L$-Lip on $[0,1]^n$ & $\Pr[|f-\operatorname{Med}f|>t]\le 4e^{-t^2/4L^2}$\\
LIS / combinatorial opt.\ & $f=$ longest incr.\ subseq., etc.\ & $\sqrt{\,}$-scale fluctuation, beats bounded diff.'s $n$-scale\\
Largest singular value & random matrix          & $\sigma_{\max}$ concentrates around its median\\
\bottomrule
\end{tabularx}

\block{Main results \& proof skeletons}

\begin{theorem}[Talagrand's convex-distance inequality]\label{thm:talagrand}
$X=(X_1,\dots,X_n)$ a product measure, $A$ measurable. Then $\E\big[e^{d_T(X,A)^2/4}\big]\le 1/\Pr[A]$.
\end{theorem}
\begin{proof}
Skeleton (induction on $n$, the hardest step in the book): \emph{(i)} write $d_T$ via convex duality (the distance to
the convex hull $U_A(x)=\operatorname{conv}\{\mathbf 1[x_i\ne y_i]:y\in A\}$); \emph{(ii)} condition on the last coordinate
and apply the inductive hypothesis to the two cases ``in $A$ / projection in $A$''; \emph{(iii)} a one-dimensional
optimization $\inf_{0\le r\le1}r^{\dots}\cdots$ merges the two cases. \TODO{fill the three-step induction; the one-dim optimization lemma $\inf_r e^{(1-r)^2/4}(\cdot)^{r}$ is the crux}
\end{proof}

\begin{corollary}[Concentration of convex Lipschitz functions]\label{cor:convex-lip}
$f:[0,1]^n\to\R$ convex and $L$-Lipschitz (Euclidean), $X$ a product measure. Then
$\Pr[|f-\operatorname{Med}f|>t]\le 4e^{-t^2/4L^2}$.
\end{corollary}
\begin{proof}\TODO{use convexity to control $f(x)-\operatorname{Med}f$ by $L\cdot d_T(x,\{f\le\operatorname{Med}f\})$ (convex $\Rightarrow$ supporting hyperplane), then apply Thm~\ref{thm:talagrand}}\end{proof}

\begin{example}[Longest increasing subsequence (LIS)]\label{ex:lis}
For a random permutation, $L_n\sim 2\sqrt n$; changing one value moves it by at most $1$, but ``certifying $L_n\ge k$ needs
only $k$ positions'', so the convex distance gives fluctuation $O(n^{1/4})$, far beating bounded differences' $O(\sqrt n)$.
\TODO{fill the certifiable / self-certifying structure $\Rightarrow d_T$ bound}
\end{example}

\begin{strand}
Why it goes this way: bounded differences only knows ``each coordinate has $c_i$''; Talagrand knows ``how \emph{many}
coordinates must conspire to change the conclusion''. When dependence is sparse (few coordinates determine $f$), convex
distance replaces $\sqrt{\sum c_i^2}$ by $\sqrt{\#\text{critical coords}}$, the source of the $n\to\sqrt n\to n^{1/4}$
chain of improvements in combinatorial optimization.
\end{strand}

\loose{Another route: Marton's transportation-cost ($W_1\le\sqrt{2\,\mathrm{KL}}$) also gives convex-distance concentration, tying it back to the entropy tensorization of §\ref{sec:entropy}. Worth listing side by side.}%
\pick{tensorize}\quad Convex distance can also be seen as the scalar precursor of the spectral concentration of §\ref{sec:matrix}.
