% ===========================================================================
\section{Surprise, and its average}\label{sec:entropy}
\warmth{0}\quad\whisper{Rare events are surprising; entropy is how surprised you are on average.}

\begin{strand}
Quantify \keyword{surprise}: an event of probability $p$ carries $-\log p$ bits of it. Certain
events ($p=1$) surprise you not at all; rare events ($p\to0$) surprise you a lot; and surprises
of \emph{independent} events add. Entropy is just the \emph{expected} surprise of a source: how
many yes/no questions, on average, to pin down its value.
\end{strand}

\block{The definition}

\begin{definition}[entropy]\label{def:entropy}
For a discrete random variable $X$ on alphabet $\X$ with law $p$,
\[
  \Ent(X)\;=\;\E\big[-\log p(X)\big]\;=\;-\sum_{x\in\X} p(x)\,\fillin[2cm].
\]
It depends only on the \emph{probabilities}, not on the values $x$. In bits ($\log_2$) it is the
average number of fair-coin questions needed to identify $X$.
\end{definition}

\recall{Three sanity checks: a sure thing has $\Ent=0$; a fair coin has $\Ent=1$ bit; and $\Ent$ is maximized by the \emph{uniform} law (most uncertain), where it equals $\log\abs{\X}$ (§\ref{sec:source}).}

\begin{yourturn}
Compute (in bits):
\[
  \Ent(\text{fair coin})=\fillin[1cm],\qquad
  \Ent\big(\tfrac14,\tfrac34\big)=\fillin[1.6cm],\qquad
  \Ent(\text{uniform on } 8)=\fillin[1cm].
\]
\recall{$1$ bit; $\ -\tfrac14\log\tfrac14-\tfrac34\log\tfrac34\approx 0.811$ bits; $\ \log_2 8=3$ bits. A biased coin is \emph{less} uncertain than a fair one, so it is cheaper to describe.}
\end{yourturn}

\begin{strand}
Why $-\log p$ and not some other measure of surprise? Because it is the \emph{only} one (up to the
choice of base) that is continuous, that grows as events get rarer, and that makes the surprise of
independent events \keyword{add}: $-\log(pq)=-\log p-\log q$. Additivity over independence is the
hinge the whole subject swings on.
\end{strand}

\loose{Continuous $X$: replace the sum by an integral to get \emph{differential} entropy $h(X)=-\int f\log f$. It can be negative and is not coordinate-free, so handle with care; the clean, operational quantity there is relative entropy (§\ref{sec:toolkit}).}
