% ===========================================================================
\section{Faces of entropy}\label{sec:faces}
\warmth{0}\quad\whisper{Training losses, channel limits, the most honest prior: all are entropy, wearing hats.}

\begin{strand}
The same price tag shows up across machine learning and communication. Each face below is entropy
or one of its relatives, doing a job.
\end{strand}

\block{The keyring}
\noindent\begin{tabularx}{\linewidth}{@{}l L@{}}
\toprule
{\color{indigo}face} & \multicolumn{1}{l}{\color{indigo}the idea}\\
\midrule
\keyword{cross-entropy} loss   & training minimizes $\E[-\log q_\theta]=\Ent(p)+\KL(p\,\Vert\,q_\theta)$, i.e.\ it closes the KL gap to the data\\
channel \keyword{capacity}     & $C=\max_p\MI(X;Y)$; you can communicate reliably at any rate below $C$, and not above\\
\keyword{maximum entropy}      & the least-committed law with given constraints: fixed variance $\Rightarrow$ Gaussian, fixed mean on $\R_{\ge0}\Rightarrow$ exponential\\
\keyword{AEP} / typical set    & $\approx 2^{n\Ent}$ sequences carry almost all probability, so $n\Ent$ bits suffice\\
\keyword{Fano}                 & if you must guess $X$ from $Y$, your error is bounded below by $\Ent(X\mid Y)$\\
\bottomrule
\end{tabularx}

\begin{yourturn}
The cross-entropy you minimize when training a classifier is $\E_{p}[-\log q_\theta(X)]$.
\begin{itemize}[leftmargin=1.3em,itemsep=1pt]
  \item It splits as $\Ent(p)+\fillin[3cm]$, and only the second term depends on $\theta$.
  \item So minimizing cross-entropy is exactly minimizing \fillin[3cm] from the model to the data,
        and the floor it can reach is $\fillin[1.6cm]$.
\end{itemize}
\recall{Splits as $\Ent(p)+\KL(p\Vert q_\theta)$; training minimizes $\KL(p\Vert q_\theta)$; the floor is $\Ent(p)$, the data's own entropy, reached only when $q_\theta=p$.}
\end{yourturn}

\block{The one idea, recovered}
\begin{strand}
Read the master table again. Each row is a number of bits, and each theorem is one of two
sentences. \keyword{Converses} (Gibbs, $L\ge\Ent$, capacity, Fano) say a price cannot be undercut;
\keyword{achievabilities} (Shannon codes, the AEP, capacity-achieving schemes) say it can be met.
Description, compression, transmission, and learning are all the same act of paying entropy, and
not a bit more than you must.
\end{strand}

\loose{Where to go next: drop probability for a single string and you get \keyword{Kolmogorov complexity} (the length of the shortest program that prints it); allow lossy reconstruction and you get rate--distortion (the price of ``good enough''); both are entropy's siblings.}
